### 2024-10-11 manuscripts/08-labor-research-v6/manuscript_main.tex
edit

In this paper, I answer the following questions. Does having a Hispanic last name affect labor market outcomes? What is the effect of identifying as Hispanic on earnings? Moreover, I aim to show that comparing Hispanic Whites to non-Hispanic Whites might create an artificially higher earnings gap since the two groups differ on many observable characteristics.\footnote{By identifying as Hispanic, I refer to individuals who self-report their ethnicity as Hispanic on surveys or other data collection instruments.} \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} Others have attempted to compare how native-born White Hispanics fare to non-Hispanic Whites and foreign-born Hispanics. In \textcite{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, the authors compare the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health. \textcite{davilaChangesRelativeEarnings2008} documents many gaps in labor market outcomes between Hispanics and Whites. They attribute a big part of this gap to differences in education, experience, immigration status, and regional differences.

@@

The distribution of the four types of children is presented Table \ref{tab:mat1}. The majority of the sample (96\%) is WW children. The second biggest group is HH, which constitutes 3\% of the sample. Inter-ethnic children, WH and HW, make up 1.35\% of the sample with 90,325 observations. Even though WH and HW are only 1.35\% of the sample, I have plenty of observations to carry out an analysis. The summary statistics for the children of the four types of marriages are presented in Table \ref{tab:c&p1}. Children of WW marriages (column 1) do better on every measure while children of HH parents (column 4) do worse than other children on every measure. Children of WH (column 2) and HW (column 3) marriages fall in between WW and HH children. Moreover, the summary statistics for those that also identify as Hispanic is reported in in Table \ref{tab:c&p2}.

@@

\section{Empirical Strategy}\label{sec:emp_model}

In this section, I present two empirical strategies. The first empirical strategy estimates the effect of having a Hispanic last name on educational outcomes and earnings. The second empirical strategy estimates the effect of identifying as Hispanic on log annual earnings.

The difference in means between Hispanics and non-Hispanic Whites could result from discrimination. It can also be caused by differences in innate abilities, skills, and parental investments. While controlling for observable skill measures, I compare children of inter-ethnic marriages, HW and WH. WH and HW children are more similar in characteristics but provide employers, and the labor market, with different signals.\footnote{WH and HW children are both half White, half Hispanic.} WH children will have a non-Hispanic last name, while HW children will have a Hispanic last name. This is a method developed by \textcite{rubinstein2014pride}.

@@

\subsection{Estimating the effect of identifying as Hispanic}

In this section, I will present an estimation strategy that would allow me to capture the effect of identifying as a Hispanic.

\begin{equation} \label{eq:iden}
Y_{ist} = \beta_{1} HW_{ist} + \beta_{2} Hispanic_i + \beta_{3} HW_{ist} \cdot Hispanic_i + X_{ist} \pi + \gamma_{st} + \phi_{ist}
\end{equation}

The coefficients of interest from equation \ref{eq:iden} are $\beta_{1}$, $\beta_{2}$ and $\beta_{3}$. The coefficients $\beta_{1}$ and $\beta_{3}$ capture the effects of Spanish-sounding last names for $HW_{ist}$ that identify as Hispanic \textit{versus} those who do not.

@@

In this section, I present the results from estimating the two specifications presented in equations \ref{eq:1a} and \ref{eq:iden}. I estimate the mean educational outcomes of White, Native-born, Hispanics aged 25-40. I estimate the mean wages of White, Native-born, Hispanic men aged 25-40 who are employed full-time and full-year (FTFY) as waged and salaried workers. The results are shown in Tables \ref{tab:lastnamereg} and \ref{tab:identreg}.

@@

Second, I find that an inter-ethnic person with a Spanish-sounding last names earns less than an inter-ethnic who has a White last name. A person with a Spanish-sounding last name earns 5 percentage points less than a person with a White last name. In other words, by comparing inter-ethnic children, a person with a Hispanic last name earns 5 percentage points less than someone with a White last name. However, more than half of the earnings gaps could be explained by educational differences. When I control for education, the last name effect decreases to a statistically insignificant 1 percentage points earnings gap. I also find a significant earnings gap between those that identify as Hispanic.

@@

I provide the results of the estimation of equation \ref{eq:1a} in Table \ref{tab:lastname-ed-reg}. I estimate the mean educational outcomes of White, U.S.-born Hispanics ages 25-40. I also restrict the sample to children of HW and WH parents. The omitted group is children of WH parents. Column one in Table \ref{tab:lastname-ed-reg} is the difference in total years of education between HW children and their WH peers. Column two is the difference in the probability of having been a high school dropout. Column three is the difference in the probability of having an associate degree. Column four is the difference in the probability of having a bachelor's degree. All regressions include controls for age, parental education and income, and state-year fixed effects.

@@

\subsection{The Effect of Having a Hispanic Last Name and Hispanic Identity on Labor Market Outcomes}

I provide the results to the estimation of equation \ref{eq:1a} in Table \ref{tab:lastnamereg}. I estimate the mean wages of White, U.S.-born, Hispanic men aged 25-40 who are employed FTFY as waged and salaried workers. I also restrict the sample to children of HW and WH (omitted) parents. Column one in Table \ref{tab:lastnamereg} is the average crude earnings gap in log annual earnings between HW workers and their WH peers. In the next 4 columns, I introduce the results with controls for hours worked, state FE, year FE, age FE, Education FE, and parental background.

@@

I provide the results to the estimation of equation \ref{eq:iden} in Table \ref{tab:identreg}. The sample in this analysis includes full-time, full-year, and wage and salary men, and I control for hours worked and age and include year FE. I also restrict the sample to children of HW and WH parents. The omitted group is children of WH parents that do not identify as Hispanic. Column one in Table \ref{tab:identreg} is the results without controlling for education. An inter-ethnic $HW_{ist}$ that identifies as Hispanic \textit{versus} $HW_{ist}$ who do not---i.e. $HW_{ist} + HW_{ist} \cdot Hispanic_i$---earns 6 percentage points less than an inter-ethnic that does not, but the result could be entirely explained by educational differences.

>>>>

In this paper, I answer the following questions. Does having a Hispanic last name affect educational outcomes? Does having a Hispanic last name affect labor market outcomes? I aim to show that comparing Hispanic Whites to non-Hispanic Whites might create an artificially higher earnings gap since the two groups differ on many observable characteristics.\footnote{By identifying as Hispanic, I refer to individuals who self-report their ethnicity as Hispanic on surveys or other data collection instruments.} \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} Others have attempted to compare how native-born White Hispanics fare to non-Hispanic Whites and foreign-born Hispanics. In \textcite{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, the authors compare the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health. \textcite{davilaChangesRelativeEarnings2008} documents many gaps in labor market outcomes between Hispanics and Whites. They attribute a big part of this gap to differences in education, experience, immigration status, and regional differences.

@@

The distribution of the four types of children is presented Table \ref{tab:mat1}. The majority of the sample (96\%) is WW children. The second biggest group is HH, which constitutes 3\% of the sample. Inter-ethnic children, WH and HW, make up 1.35\% of the sample with 90,325 observations. Even though WH and HW are only 1.35\% of the sample, I have plenty of observations to carry out an analysis. The summary statistics for the children of the four types of marriages are presented in Table \ref{tab:c&p1}. Children of WW marriages (Column 1) do better on every measure while children of HH parents (Column 4) do worse than other children on every measure. Children of WH (Column 2) and HW (Column 3) marriages fall in between WW and HH children. Moreover, the summary statistics for those that also identify as Hispanic is reported in in Table \ref{tab:c&p2}.

@@

\section{Empirical Approach}\label{sec:emp_model}

In this section, I present my empirical Approach. The specification estimates the effect of having a Hispanic last name on educational outcomes and earnings. The difference in means between Hispanics and non-Hispanic Whites could result from discrimination. It can also be caused by differences in innate abilities, skills, and parental investments. While controlling for observable skill measures, I compare children of inter-ethnic marriages, HW and WH. WH and HW children are more similar in characteristics but provide employers, and the labor market, with different signals.\footnote{WH and HW children are both half White, half Hispanic.} WH children will have a non-Hispanic last name, while HW children will have a Hispanic last name. This is a method developed by \textcite{rubinstein2014pride}.

@@

@@

In this section, I present the results from estimating the specification presented in equation \ref{eq:1a}. I estimate the mean educational outcomes of White, Native-born (results in Table \ref{tab:lastname-ed-reg}), Hispanics aged 25-40, and mean unemployment of White, Native-born, Hispanics men aged 25-40 (results in Table \ref{tab:lastnamereg-emp}). I estimate the mean wages of White, Native-born, Hispanic men aged 25-40 who are employed full-time and full-year (FTFY) as waged and salaried workers (results in Table \ref{tab:lastnamereg}).

@@

Second, I find that an inter-ethnic person with a Spanish-sounding last names are 1 percentage point more likely to be unemployed and earn less than an inter-ethnic who has a White last name. A person with a Spanish-sounding last name earns 5 percentage points less than a person with a White last name. In other words, by comparing inter-ethnic children, a person with a Hispanic last name earns 5 percentage points less than someone with a White last name. However, more than half of the earnings gaps could be explained by educational differences. When I control for education, the last name effect decreases to a statistically insignificant 1 percentage points earnings gap. I also find a significant earnings gap between those that identify as Hispanic.

@@

I provide the results of the estimation of equation \ref{eq:1a} in Table \ref{tab:lastname-ed-reg}. I estimate the mean educational outcomes of White, U.S.-born Hispanics ages 25-40. I also restrict the sample to children of HW and WH parents. The omitted group is children of WH parents. Column 1 in Table \ref{tab:lastname-ed-reg} is the difference in total years of education between HW children and their WH peers. Column 2 is the difference in the probability of having been a high school dropout. Column 3 is the difference in the probability of having an associate degree. Column 4 is the difference in the probability of having a bachelor's degree. All regressions include controls for age, parental education and income, and state-year fixed effects.

@@

\subsection{The Effect of Having a Hispanic Last Name on Labor Market Outcomes}

I provide the results to the estimation of equation \ref{eq:1a} in Tables \ref{tab:lastnamereg-emp} and \ref{tab:lastnamereg} on unemployment and log earnings. I estimate the mean unemployment of White, Native-born, Hispanics men aged 25-40. I estimate the mean wages of White, U.S.-born, Hispanic men aged 25-40 who are employed FTFY as waged and salaried workers. I also restrict the sample to children of HW and WH (omitted) parents. Column 1 in Table \ref{tab:lastnamereg} is the average crude earnings gap in log annual earnings between HW workers and their WH peers. In the next 4 columns, I introduce the results with controls for hours worked, state FE, year FE, age FE, Education FE, and parental background.

In addition to examining earnings, I also analyzed the effect of having a Hispanic last name on unemployment rates. Table \ref{tab:lastnamereg-emp} presents the results of this analysis. Column 1 shows that individuals with Hispanic last names (HW) have a 1 percentage point higher unemployment rate compared to those with White last names. This gap persists even after controlling for age, state fixed effects, year fixed effects, and state-year fixed effects (Column 2), though the significance level drops. When education is included as a control (Column 3), the gap remains at 1 percentage point. Finally, after controlling for parental background (Column 4), the 1 percentage point difference in unemployment rates becomes statistically insignificant. These results suggest that while there is an initial unemployment gap associated with having a Hispanic last name, much of this difference can be explained by factors such as education and parental background. The mean unemployment rate for individuals with Hispanic last names (HW) is 7\% across all specifications.

@@

I provide the results to the estimation of equation \ref{eq:iden} in Table \ref{tab:identreg}. The sample in this analysis includes full-time, full-year, and wage and salary men, and I control for hours worked and age and include year FE. I also restrict the sample to children of HW and WH parents. The omitted group is children of WH parents that do not identify as Hispanic. Column 1 in Table \ref{tab:identreg} is the results without controlling for education. An inter-ethnic $HW_{ist}$ that identifies as Hispanic \textit{versus} $HW_{ist}$ who do not---i.e. $HW_{ist} + HW_{ist} \cdot Hispanic_i$---earns 6 percentage points less than an inter-ethnic that does not, but the result could be entirely explained by educational differences.

### 2024-10-11 manuscripts/08-labor-research-v6/manuscript_title_etc.tex
edit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 2. Define Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newcommand{\PAPERTITLE}{\href{https://hhadah.github.io/hispanic-last-names/my_paper/Hadah-last-names-draft.pdf}{The Impact of Hispanic Last Names and Identity on Labor Market Outcomes}}
\newcommand{\PAPERTITLE}{The Impact of Hispanic Last Names and Identity on Educational and Labor Market Outcomes}

@@

Do individuals with Hispanic names face labor market discrimination? This study analyzes the impact of Hispanic-sounding surnames on wages among inter-ethnic children with one White and one Hispanic parent. I find that individuals with Hispanic surnames often earn less, with a notable wage gap favoring those with White surnames. People born to Hispanic fathers and White mothers receive 0.3 years of education less than those born to White fathers and Hispanic mothers. Males born to Hispanic fathers and White mothers earn 5 percentage points less than those born to White fathers and Hispanic mothers, largely due to educational differences. Additionally, men with Spanish-sounding last names who identify as Hispanic earn significantly less, again due to educational disparities. \PAPERJEL}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 2. Define Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newcommand{\PAPERTITLE}{\href{https://hhadah.github.io/hispanic-last-names/my_paper/Hadah-last-names-draft.pdf}{The Impact of Hispanic Last Names and Identity on Labor Market Outcomes}}
\newcommand{\PAPERTITLE}{The Impact of Hispanic Last Names on Educational and Labor Market Outcomes}

@@

Do individuals with Hispanic names face labor market discrimination? This study analyzes the impact of Hispanic-sounding surnames on wages among inter-ethnic children with one White and one Hispanic parent. I find that individuals with Hispanic surnames often earn less, with a notable wage gap favoring those with White surnames. People born to Hispanic fathers and White mothers receive 0.3 years of education less than those born to White fathers and Hispanic mothers. Males born to Hispanic fathers and White mothers are 1 percentage point more likely to be unemployed, and they also earn 5 percentage points less than those born to White fathers and Hispanic mothers. The earnings gap is largely due to educational differences. \PAPERJEL}
